Verifying Spatial Properties of Array Computations

### 2017-04-18 oopsla17/algorithms.tex
second part of inference

The preceding source code analysis converts the concrete syntax of
array subscripts into a sets of subscript schemes. Inference then has
two parts. First, ``adjacent'' index schemes are coalesced into a
smaller sets of index schemes, remaining in union normal
form. Secondly, the resulting union normal form is converted to the
specification syntax.

@@

\newcommand{\contig}[2]{\mathit{adjacent}(#1, #2)}
\begin{defn}[Adjacent]
  \label{def:contiguity}
  Two index schemes $S$ and $T$ are \emph{adjacent} written
  $\contig{S}{T}$, iff $\pi_i(S) = \pi_i(T)$ for all $1 \leq i \leq N$
  apart from some dimension $k$ such that $\pi_k(S) = [a, b]$ and
  $\pi_k(T) = [b+1, c]$.
\end{defn}
%
Given a set of indexing scheme $\mathcal{S}$ and a particular index scheme $S$
we generate the set of coalescing of $S$ with adjacent index schemes:
%
\begin{equation*}
\mathit{coalesce}(S, \mathcal{S}) = 
  \{\,S \cup T \mid T \in \mathcal{S} \, \wedge \,
  \contig{S}{T} \}
\end{equation*}
%
This is then used by the following recursive procedure:
%
\begin{align*}
  \mathit{coalesceStep}(\mathcal{S}) =
  \exists T \in \mathcal{S} . 
  \begin{cases}
    \{T\} \cup \mathit{coalesceStep}(\mathcal{S} - \{T\}) &
    \mathit{coalesce}(T, \mathcal{S}) = \emptyset \\
    \mathit{coalesceStep}(\mathit{coalesce}(T, \mathcal{S}) \cup
                                        \mathcal{S} - \{T\}) & \textit{otherwise}
  \end{cases}
\end{align*}
%
This give a specification rather than an implementation. Our
implementation represents sets by a list, and so
$\exists T \in \mathcal{S}$ above corresponds to deconstructing
the list into its head element $S$ (picking an element from the set).
If $T$ has no adjacent index schemes, then coalescing is attempted on
the rest of the set, and $T$ is returned in the result. Otherwise,
the set of coalesced index scheme is computed and this is passed to
a recursive procedure called to $\mathit{coalesceStep}$, along with the rest of
the elements).

The fixed-point of $\textit{coalesceStep}(\mathcal{S})$ is computed to
give a covering over the initial subscript space.
%
\begin{restatable}{lem}{closureInference}
  \label{lem:closuer-inference}
  For a set of index schemes $\mathcal{S}$, then
  $\mathit{coalesceSet}(\mathcal{S})$ is well-defined and is a set of
  index schemes in Union Normal Form.
\end{restatable}
%
For our example, the fixed-pointed of $\mathit{coalesceStep}$ is reached within two
steps:
%
\begin{align*}
  \mathit{coalesceStep}(\mathcal{S}_0) & =
  \{[-1,0] \times [0,0],\,[0,0] \times [-1, 0],\,[0,0] \times
    [0,1],\,[0,1] \times [0,0]\} \\
  \mathit{coalesceStep}^2(\mathcal{S}_0) & =
  \{[-1, 1] \times [0, 0],\,[0, 0] \times [-1, 1]\} = \mathit{coalesceStep}^3(\mathcal{S}_0)
\end{align*}
%
%\todo{Prove that this is in $Region_N$}

@@

\newcommand{\finalSet}{\mathcal{S}_\omega}
Let the covering indexing scheme from the fixed point of
$\mathit{coalesceStep}$ on input $\mathcal{S}_0$ be written as
$\finalSet{}$.  Next, $\finalSet{}$ the is translated into specification
syntax. This happens in four stages.

\begin{enumerate}[leftmargin=1.5em]
  \item We check whether $\finalSet{}$
  is the top indexing scheme $\bz{}^N$. This occurs if absolute indices
  occur in each of the dimensions in any of the array subscripts,
  leading to an unconstrained access pattern, which is not represented
  in our syntax, ending inference.

  \item Otherwise, we determine if the indexing schemes are interval
  schemes so that they can be represented exactly. It not, they are
  altered into interval schemes to be represented as approximations.

  \item We split the interval schemes so that the region constants can
  express them.

  \item Finally, we convert the interval schemes into a joins of meets
  of region constants.
\end{enumerate}

If there is all infinite indexing scheme, then we cannot represent it
syntactically. This happens when use of constant indices in one or more array
terms causes the offsets to be subsumed. The simplest case is when there is an
array term with all of its indices constant.

In our example, there are no constant indices, so we proceed.

After handling the trivial case, correct approximation is chosen. Recall
\Cref{def:interval-scheme}, an indexing scheme is an interval scheme if each of
the sets in products are holed intervals. If that is the case, we trivially
convert the sets in each dimension of indexing schemes to holed interval
representation and proceed to the next stage and output an exact specification
at the end. If not, then the output will definitely have an upper bound and
potentially a lower bound.

Upper bound is established by \emph{elongating} index schemes in multiple
dimensions such that they become interval schemes. The elongation function is
given as
%
\begin{align*}
  \mathit{elongate}([a,b]) = \begin{cases}
    \interv{0}{b}{\textsf{false}} & a > 1 \\
    \interv{a}{0}{\textsf{false}} & b < -1
  \end{cases}
\end{align*}

Since index schemes of the previous stage are all contiguous and form closed
intervals, 0 lying between the lower and upper bounds of intervals is sufficient
to form an interval scheme. If any dimension of an index scheme is elongated,
then the whole index scheme is elongated. In the initial set of index schemes,
all schemes that were already interval schemes form a lower bound, while those
schemes and the new interval schemes generated by elongation form the upper
bound. The reason a lower bound may not be generated is all indexing schemes
with might need elongation.

The index schemes in the example are all interval schemes, so we proceed to the
next stage noting the output will be exact.

There is no direct translation from an interval scheme such as $\interv{-2}{1}{}
\times \interv{-1}{1}{}$ to region constants. The reason is region constants
\texttt{forward} and \texttt{backward} require upper and lower bounds to be 0
respectively and \texttt{centered} requires the upper bound to be the negation
of the lower bound. The solution is to use \Cref{lem:vector-union} to break up
interval schemes into two producing $\interv{-2}{0}{} \times \interv{-1}{1}{}$
and $\interv{0}{1}{} \times \interv{-1}{1}{}$.

The extended example does not have such representation problem, so the interval
schemes are left as they are.

The final stage of inference is simply mapping interval schemes into region
constants. Each interval scheme is connected with \texttt{+}. Individual schemes
are decomposed using \Cref{lem:vector-intersect} such that each of the resulting
index schemes have all but one dimensions set to \interv{-\infty}{\infty}{}.
Each of these interval schemes are mapped to region constants using
\Cref{subfig:region-model} and connected with \texttt{*}.

In the example, $\interv{-1}{1}{} \times \interv{0}{0}{}$ is decomposed into
$\interv{-1}{1}{} \times \interv{-\infty}{\infty}{}$ and
$\interv{-\infty}{\infty}{} \times \interv{0}{0}{}$. These are then mapped to
\texttt{\stencil{c}{1}{1}{} * \stencil{p}{2}{}{}}. Decomposition of
$\interv{0}{0}{} \times \interv{-1}{1}{}$ is similar. When combined we obtain the
following specification:
%
\begin{equation*}
  \texttt{stencil readOnce,} \;
  \stencil[s]{c}{1}{1}{} \texttt{*} \stencil[s]{p}{2}{}{} \; \texttt{+} \;
  \stencil[s]{p}{1}{}{}  \texttt{*} \stencil[s]{c}{2}{1}{}
\end{equation*}

>>>>

The program analysis converts the concrete syntax of
array subscripts into a sets of subscript schemes. Inference then has
two parts. First, ``adjacent'' index schemes are coalesced into a
smaller sets of index schemes, remaining in union normal
form. Secondly, the resulting union normal form is converted to the
specification syntax.

@@

\newcommand{\contig}[2]{\mathit{adjacent}(#1, #2)}
\begin{defn}[Adjacent]
  \label{def:contiguity}
  Two index schemes $S$ and $T$ are \emph{adjacent} written
  $\contig{S}{T}$, iff $\pi_i(S) = \pi_i(T)$ for all $1 \leq i \leq N$
  apart from some dimension $k$ such that $\pi_k(S) = [a, b]$ and
  $\pi_k(T) = [b+1, c]$.
\end{defn}
%
Given a set of indexing scheme $\mathcal{S}$ and a particular index scheme $S$
we generate the set of coalescing of $S$ with adjacent index schemes:
%
\begin{equation*}
\mathit{coalesce}(S, \mathcal{S}) = 
  \{\,S \cup T \mid T \in \mathcal{S} \, \wedge \,
  \contig{S}{T} \}
\end{equation*}
%
This is then used by the following recursive procedure:
%
\begin{align*}
  \mathit{coalesceStep}(\mathcal{S}) =
  \exists T \in \mathcal{S} . 
  \begin{cases}
    \{T\} \cup \mathit{coalesceStep}(\mathcal{S} - \{T\}) &
    \mathit{coalesce}(T, \mathcal{S}) = \emptyset \\
    \mathit{coalesceStep}(\mathit{coalesce}(T, \mathcal{S}) \cup
                                        \mathcal{S} - \{T\}) & \textit{otherwise}
  \end{cases}
\end{align*}
%
This give a specification rather than an implementation. Our
implementation represents sets by a list, and so
$\exists T \in \mathcal{S}$ above corresponds to deconstructing
the list into its head element $S$ (picking an element from the set).
If $T$ has no adjacent index schemes, then coalescing is attempted on
the rest of the set, and $T$ is returned in the result. Otherwise,
the set of coalesced index scheme is computed and this is passed to
a recursive called to $\mathit{coalesceStep}$, along with the rest of
the elements).

The fixed-point of $\textit{coalesceStep}(\mathcal{S})$ is computed to
give a covering over the initial subscript space.
%
\begin{restatable}{lem}{closureInference}
  \label{lem:closuer-inference}
  For a set of index schemes $\mathcal{S}$, then
  $\mathit{coalesceSet}(\mathcal{S})$ is a set of
  index schemes in Union Normal Form.
\end{restatable}
%
For our example, the fixed-pointed of $\mathit{coalesceStep}$ is reached within two
steps:
%
\begin{align*}
  \mathit{coalesceStep}(\mathcal{S}_0) & =
  \{[-1,0] \times [0,0],\,[0,0] \times [-1, 0],\,[0,0] \times
    [0,1],\,[0,1] \times [0,0]\} \\
  \mathit{coalesceStep}^2(\mathcal{S}_0) & =
  \{[-1, 1] \times [0, 0],\,[0, 0] \times [-1, 1]\} = \mathit{coalesceStep}^3(\mathcal{S}_0)
\end{align*}
%
%\todo{Prove that this is in $Region_N$}

@@

\newcommand{\finalSet}{\mathcal{S}_\omega}
Let the covering indexing scheme from the fixed point of
$\mathit{coalesceStep}$ on input $\mathcal{S}_0$ be written as
$\finalSet{}$.  Next, $\finalSet{}$ is translated into specification
syntax. This happens in three stages.

\begin{enumerate}[leftmargin=2em]
  \item We check whether $\finalSet{}$
  is the top indexing scheme $\bz{}^N$. This occurs if absolute indices
  appear in each of the dimension across the array subscripts,
  leading to an unconstrained access pattern, which is not represented
  in our syntax, ending inference.

  \item Otherwise, we determine if the indexing schemes are interval
  schemes (\Cref{def:interval-scheme}) so that they can be represented
  exactly by our syntax. It not, they are altered into interval
  schemes to be represented as approximations.

  \item Interval schemes into a joins of meets
  of region constants, where some intervals are split into separate
  region constants.
\end{enumerate}
%
An index scheme is an interval scheme (\Cref{def:interval-scheme}) if
each of vector can be represented as a vector of holed intervals
$\interv{a}{b}{c}$.  If so, we convert each index scheme to an
interval scheme as a vector of holed interval representation.

Since index schemes of the previous stage are all contiguous and form
closed intervals, a lower bound $\leq 0$ and an upper bound $\geq 0$
is sufficient to form an interval scheme.  Otherwise, an approximation
specification is generated. An upper bound is established by
\emph{elongating} any index schemes in multiple dimensions such that
they become interval schemes. This is given by the function:
%
\begin{align*}
  \mathit{elongate}([a,b]) = \begin{cases}
    \interv{0}{b}{\textsf{false}} & a > 1 \\
    \interv{a}{0}{\textsf{false}} & b < -1
  \end{cases}
\end{align*}
%
If any dimension of an index scheme is elongated, then the whole index
scheme is elongated. In $\finalSet$, all index schemes that are
representable as interval schemes form a lower bound, whilst those
that do not produce an upper bound by their elongation. A lower bound
may not be generated if all indexing schemes are not interval schemes.
In the case of our example, the resulting $\finalSet$ comprises a set
of interval schemes, thus we can generate an exact specification.

\paragraph{Translation from interval schemes}

Recall the model of region constants in \Cref{subfig:region-model}
where $\interp{-}$ maps to holed intervals which have either $0$ as
the lower-bound or upper-bound for \fortran{forward} and
\fortran{backward} respectively, or both $0$ for \fortran{pointed}, or
$-k$ and $k$ lower and upper bounds for \fortran{centered}. We can
invert this map to generate region constants from holed
intervals. However, some intervals computed from the above steps
may not the constraints of this interpretation,
\eg{}, $\interv{-2}{1}{} \times \interv{-1}{1}{}$.  By
\Cref{lem:vector-union}, we can split interval schemes into two into
the relevant form, for example, producing
$\interv{-2}{0}{} \times \interv{-1}{1}{}$ and
$\interv{0}{1}{} \times \interv{-1}{1}{}$. 
The following function
$\textit{split}$ iterates over the components of an index scheme,
splitting apart interval schemes when necessary:
%
\begin{align*}
  %\textit{split}(\emptyset) = \emptyset \quad
  \textit{split}(\interv{a}{b}{c} \times S) =
   \begin{cases}
  \{ \interv{a}{0}{c} \times T \mid T \in \textit{split}(S) \}
  \cup
  \{ \interv{0}{b}{c} \times T \mid T \in \textit{split}(S) \}
  & a < 0 \vee b > 0 \vee (|a| \neq b) \\
  \{ \interv{a}{b}{c} \times T \mid T \in \textit{split}(S) \}
\end{cases}
\end{align*}
%
This forms a set of interval schemes to which $\interp{-}^{-1}$ can be
applied. In the case of our example, no splitting is necessary.

The final stage of inference is simply mapping interval schemes into
region constants and combining these. Each translated interval scheme
is connected with \texttt{+}. Individual schemes are decomposed using
\Cref{lem:vector-intersect} such that each of the resulting index
schemes have all but one dimensions set to \interv{-\infty}{\infty}{}.
Each of these interval schemes are mapped to region constants using
\Cref{subfig:region-model} and connected with \texttt{*}.
%
\begin{align*}
\textit{convert}_N(\mathcal{S}) = \sum_{S \in S} \prod_{1 \leq i \leq N} \interp{\pi_i(S)}^{-1}
\end{align*}
%
In the example, $\interv{-1}{1}{} \times \interv{0}{0}{}$ is decomposed into
$\interv{-1}{1}{} \times \interv{-\infty}{\infty}{}$ and
$\interv{-\infty}{\infty}{} \times \interv{0}{0}{}$. These are then mapped to
\texttt{\stencil{c}{1}{1}{} * \stencil{p}{2}{}{}}. Decomposition of
$\interv{0}{0}{} \times \interv{-1}{1}{}$ is similar. When combined we
obtain the following specification:
%
\begin{equation*}
  \texttt{stencil readOnce,} \;
  \stencil[s]{c}{1}{1}{} \texttt{*} \stencil[s]{p}{2}{}{} \; \texttt{+} \;
  \stencil[s]{p}{1}{}{}  \texttt{*} \stencil[s]{c}{2}{1}{}
\end{equation*}

### 2017-04-18 oopsla17/introduction.tex
second part of inference

Such array computations are prone to error in their indexing
terms. For example, a logical off-by-one-error might manifest itself
as writing $\texttt{a(i)}$ instead of $\texttt{a(i-1)}$ (we revisit an
example we found of this in \Cref{sec:case-studies}).
Errors also arise by simple lexical mistakes when large amounts of
fine-grained indexing are involved in a single expression. For
example, the following snippet from a Navier-Stokes
fluid model~\cite{griebel1997numerical} has two arrays which are
read with different data access patterns, across two dimensions, with
dense index-manipulation code:
%%
\begin{minted}[firstnumber=20,xleftmargin=2em]{fortran}
du2dx = ((u(i,j)+u(i+1,j))*(u(i,j)+u(i+1,j))+gamma*abs(u(i,j)+u(i+1,j))*(u(i,j)-u(i+1,j))- &
  (u(i-1,j)+u(i,j))*(u(i-1,j)+u(i,j))-gamma*abs(u(i-1,j)+u(i,j))*(u(i-1,j)-u(i,j))) /(4.0*delx)

>>>>

Such array computations are prone to error in their indexing
terms. For example, a logical off-by-one-error might manifest itself
as writing $\texttt{a(i)}$ instead of $\texttt{a(i-1)}$ (we revisit an
example we found of this in \Cref{sec:case-studies}).
Errors also arise by simple lexical mistakes when large amounts of
fine-grained indexing are involved in a single expression. For
example, the following snippet from a Navier-Stokes
fluid model~\cite{griebel1997numerical} has two arrays which are
read with different data access patterns, across two dimensions, with
dense index-manipulation:
%%
\begin{minted}[firstnumber=20,xleftmargin=2em]{fortran}
du2dx = ((u(i,j)+u(i+1,j))*(u(i,j)+u(i+1,j))+gamma*abs(u(i,j)+u(i+1,j))*(u(i,j)-u(i+1,j))- &
  (u(i-1,j)+u(i,j))*(u(i-1,j)+u(i,j))-gamma*abs(u(i-1,j)+u(i,j))*(u(i-1,j)-u(i,j))) /(4.0*delx)

### 2017-04-18 oopsla17/lang.tex
second part of inference

The region operators \term{+} and \term{*} respectively combine
regions by union and intersection. The intersection of two regions
$r \term{*} s$ means that any indices in the specified code must be
consistent with both $r$ and $s$ simultaneously.
 Dually, for the union of two regions
 $r \term{+} s$ means that indices in the specified code must be
 inconsistent with one of $r$ or $s$, or both.
For example, the following \emph{nine-point stencil}
has a specification given by the product of two \texttt{centered}
regions in each dimension:
%%
\begin{minted}[breakindent=2.9em]{fortran}
x = a(i, j)   + a(i-1, j)   + a(i+1, j)
y = a(i, j-1) + a(i-1, j-1) + a(i+1, j-1)
z = a(i, j+1) + a(i-1, j+1) + a(i+1, j+1)
!= stencil centered(depth=1, dim=1) * centered(depth=1, dim=2) :: a
b(i, j) = (x + y + z) / 9.0
\end{minted}
%
%This pattern is common in image convolution applications.
The specification ranges over the
values that flow to the array subscript on the left-hand side,
and so ranges over the intermediate assignments to \term{x},
\term{y}, and \term{z}. Each index in the code is consistent
with both specifications simultaneously, \eg{}, \texttt{a(i-1, j+1)}
is within the centered region in dimension $1$ and the centered region
in dimension $2$.

>>>>

The region operators \term{+} and \term{*} respectively combine
regions by union and intersection. The intersection of two regions
$r \term{*} s$ means that any indices in the specified code must be
consistent with both $r$ and $s$ simultaneously.
 Dually, for the union of two regions
 $r \term{+} s$ means that indices in the specified code must be
 inconsistent with one of $r$ or $s$, or both.
For example, the following \emph{nine-point stencil}
has a specification given by the product of two \texttt{centered}
regions in each dimension:
%%
\begin{minted}[breakindent=2.9em]{fortran}
x = a(i, j)   + a(i-1, j)   + a(i+1, j)
y = a(i, j-1) + a(i-1, j-1) + a(i+1, j-1)
z = a(i, j+1) + a(i-1, j+1) + a(i+1, j+1)
!= stencil centered(depth=1, dim=1) * centered(depth=1, dim=2) :: a
b(i, j) = (x + y + z) / 9.0
\end{minted}
%
%This pattern is common in image convolution applications.
The specification ranges over the
values that flow to the array subscript on the left-hand side,
and so ranges over the assignments to \term{x},
\term{y}, and \term{z}. Each index in the code is consistent
with both parts of the specification, \eg{}, \texttt{a(i-1, j+1)}
is within the centered region in dimension $1$ and the centered region
in dimension $2$.
